### 2021-06-17 submission/manuscript.Rmd
Additional edits

## Importance
Diarrheal samples from patients taking laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are similarities between the bacterial communities from people with diarrhea or *C. difficile* infections (CDI) including lower diversity compared to communities from healthy patients, which led us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge. Our findings suggest that osmotic laxatives disrupt colonization resistance to *C. difficile*, and prevent clearance among mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients are separated into two groups based on stool consistency, there are similar microbiota features between samples from CDI patients and *C. difficile* negative patients with diarrhea compared to samples that were *C. difficile* negative with non-diarrheal stool consistency [@Schubert2014]. The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa had higher relative abundances in communities from CDI patients. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Because time and treatment group influenced most of the variation between communities, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, the Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin alone through thirty days post-challenge (Fig. 2C; Data Set S1, sheet 2). We next identified the bacterial genera whose relative abundances shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera whose relative abundances were altered by PEG treatment (Data Set S1, sheet 3). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

@@

One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment. In fact, it has been hypothesized that *C. difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C. difficile* biofilms in the mucus layer were recently identified in patients as aggregates with *Fusobacterium nucleatum* [@Engevik2021]. There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C. difficile* in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. *Bacteroides* has the potential to degrade mucus and the osmotic laxative may have allowed *Bacteroides* to colonize the mucosal niche by degrading mucin glycans with glycosyl hydrolases that are absent in *C. difficile* [@Engevik2020]. *Bacteroides* persistent in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B. fragilis* and *E. coli* has previously been described [@Dejea2018].  A separate study demonstrated *C. difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015]. However, protective roles for *Bacteroides* have also been demonstrated. For example, *B. fragilis* prevented CDI morbidity in a mouse model and inhibited *C. difficile* adherence *in vitro* [@Deng2018]. In coculture experiments *B. longum* decreased *C. difficile* biofilm formation while *B. thetaiotamicron* enhanced biofilm formation [@Normington2021] and *B. dorei* reduced *C. difficile* growth in a 9-species community *in vitro* [@Hassall2021]. Therefore, whether *Bacteroides* is detrimental or beneficial in the context of *C. difficile* infection or colonization is still unclear, but the niche and interactions with other bacteria may contribute.

@@

Despite identifying an altered compositional profile that included higher relative abundance of the *C. difficile* sequence in the mucosal tissues of mice treated with 5-day PEG compared to the clindamycin group, we did not see a difference in histopathology scores between the groups. One reason there was no difference could be the *C. difficile* strain used, *C. difficile* 630 results in mild histopathology summary scores in mice compared to VPI 10463 despite both strains producing toxin in mice [@Theriot2011]. Part of our hypothesis for why there could have been increased histopathology scores in PEG-treated mice was because PEG was previously shown to disrupt the mucus layer in mice. However, recent studies demonstrated that broad spectrum antibiotics can also disrupt the host mucosal barrier in mice [@Kester2020; @Bergstrom2020]. Future research is needed to tease out the interplay between medications that influence the mucus layer and different strains of *C. difficile* in the context of CDIs.

It is more difficult to interpret what are findings mean in the context of *C. difficile* colonization resistance in human patients. The main difficulty being that most hospitals recommend not performing *C. difficile* testing if the patient is currently taking a laxative. This recommendation is in accordance with the Infectious Diseases Society of America and Society for Healthcare Epidemiology of America guidelines [@McDonald2018b]. The rationale behind the recommendation is that patients taking laxatives may be asymptomatically colonized with *C. difficile*, resulting in unnecessary antibiotic treatment [@Ahmad2017; @Bilinskaya2018; @Cook2020]. Furthermore, some studies identified laxatives as a risk factor for developing CDIs or recurrent CDIs [@Appaneal2019; @Dubberke2011; @Obritsch2010] and a recent study found the proportion of severe CDIs was similar between patients taking and not taking laxatives [@White2019]. However, there have also been some studies that suggest laxatives are not a risk factor for developing CDIs [@Poirier2019; @Huang2014]. Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014]. Further studies to examine the relationship between laxatives, *C. difficile* colonization, and CDIs are warranted.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters of *C. difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate casettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.

>>>>

## Importance
Diarrheal samples from patients taking laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are similarities between the bacterial communities from people with diarrhea or *C. difficile* infections (CDI) including lower diversity compared to communities from healthy patients. This observation led us to hypothesize that diarrhea may be an indicator of *C. difficile* susceptibility. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge. Our findings suggest that osmotic laxatives disrupt colonization resistance to *C. difficile*, and prevent clearance among mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients were separated into two groups based on stool consistency, there were similar microbiota features between samples from CDI patients and *C. difficile* negative patients with diarrhea compared to non-diarrheal samples that were *C. difficile* negative [@Schubert2014]. The similar community features between CDI patients and patients with diarrhea included low alpha diversity and there were also only 6 bacterial taxa had higher relative abundances in communities from CDI patients. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

Depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Because time and treatment group influenced most of the variation between communities, we next explored whether there were differences in community diversity and composition between treatment groups. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, the Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin alone through thirty days post-challenge (Fig. 2C; Data Set S1, sheet 2). We next identified the bacterial genera whose relative abundances shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera whose relative abundances were altered by PEG treatment (Data Set S1, sheet 3). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

@@

One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment. In fact, it has been hypothesized that *C. difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C. difficile* biofilms in the mucus layer were recently identified in patients as aggregates with *Fusobacterium nucleatum* [@Engevik2021]. There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C. difficile* in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. *Bacteroides* has the potential to degrade mucus and the osmotic laxative may have allowed *Bacteroides* to colonize the mucosal niche by degrading mucin glycans with glycosyl hydrolases that are absent in *C. difficile* [@Engevik2020]. *Bacteroides* persistence in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B. fragilis* and *E. coli* has previously been described [@Dejea2018]. A separate study demonstrated *C. difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015]. However, protective roles for *Bacteroides* have also been demonstrated. For example, *B. fragilis* prevented CDI morbidity in a mouse model and inhibited *C. difficile* adherence *in vitro* [@Deng2018]. In coculture experiments *B. longum* decreased *C. difficile* biofilm formation while *B. thetaiotamicron* enhanced biofilm formation [@Normington2021] and *B. dorei* reduced *C. difficile* growth in a 9-species community *in vitro* [@Hassall2021]. Therefore, whether *Bacteroides* is detrimental or beneficial in the context of *C. difficile* infection or colonization is still unclear, but the niche and interactions with other bacteria may contribute.

@@

Despite identifying an altered compositional profile that included higher relative abundance of the *C. difficile* sequence in the mucosal tissues of mice treated with 5-day PEG compared to the clindamycin group, we did not see a difference in histopathology scores between the groups. One reason there was no difference could be the *C. difficile* strain used, *C. difficile* 630 results in mild histopathology summary scores in mice compared to VPI 10463 despite both strains producing toxin in mice [@Theriot2011]. Part of our hypothesis for why there could have been increased histopathology scores in PEG-treated mice was because PEG was previously shown to disrupt the mucus layer in mice. However, recent studies demonstrated that broad spectrum antibiotics can also disrupt the host mucosal barrier in mice [@Kester2020; @Bergstrom2020]. Further research is needed to tease out the interplay between medications that influence the mucus layer and different strains of *C. difficile* in the context of CDIs.

It is more difficult to interpret what are findings mean in the context of *C. difficile* colonization resistance in human patients. The main difficulty being that most hospitals recommend not performing *C. difficile* testing if the patient is currently taking a laxative. This recommendation is in accordance with the Infectious Diseases Society of America and Society for Healthcare Epidemiology of America guidelines [@McDonald2018b]. The rationale behind the recommendation is that patients taking laxatives may be asymptomatically colonized with *C. difficile*, resulting in unnecessary antibiotic treatment [@Ahmad2017; @Bilinskaya2018; @Cook2020]. Furthermore, some studies identified laxatives as a risk factor for developing CDIs or recurrent CDIs [@Appaneal2019; @Dubberke2011; @Obritsch2010] and a recent study found the proportion of severe CDIs was similar between patients taking and not taking laxatives [@White2019]. However, there have also been some studies that suggest laxatives are not a risk factor for developing CDIs [@Poirier2019; @Huang2014]. Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014]. Additional studies to examine the relationship between laxatives, *C. difficile* colonization, and CDIs are warranted.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters containing 10^5^ *C. difficile* 630 spores, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate cassettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.
